---
title: "Clinical malaria trends"
subtitle: "2002-2016"
author: "Bea and Joe"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
library(cism)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, 
                      cache.extra = packageVersion('tufte'),
                      echo = FALSE)
options(htmltools.dir.version = FALSE)
```



# Analysis

## OPD

### Absolute number of malaria cases over time

```{r}
x <- opd %>%
  mutate(month=format(date, "%m"),
         year=format(date, "%Y")) %>%
  group_by(year, month) %>%
  summarise(totvis=n(),
            malcase=length(which(malaria))) %>%
  mutate(malprev=malcase/totvis) %>%
  mutate(dumdate=as.Date(paste0(year, "-", month, "-15")))

library(ggthemes)  
library(cism)
ggplot(data=x, 
       aes(x=dumdate, y=malcase)) +
  geom_area(aes(x=dumdate, y=totvis), color="blue", fill="blue") +
  geom_line(fill="red") +
  labs(title="OPD Visits and Malarias", x="Date", y="Cases") +
  theme_economist()
  
```

#### By age group

Age groups should be divided differently than previously^[<1, 1-3, 3-5, 5-10, 10-15]

### Malaria cases over time (relative to all OPD cases)

#### By age group

### Incidence: Malaria cases over time (relative to population at risk)

### Clinical profile of malaria

*What does non-complicated malaria look like over time?*

- Parasite densitites^[We still need servolab stuff]
- Body temperature
- Hemoglobin
- Age distributions over time

#### Age group analysis

Cohort analysis with time-standardized x axis and cohort colored lines

(Everything up to here is for Katarina's paper)

### Hotspot identifical / Spatial scan statistic

Bayesian spatial-temporal analysis

Validate using actual locations.

Compare clinical hotspots from OPD with community hotspots from xmal.



## INPD trends

### Absolute cases over time

### Cases as percentage of all INPD cases over time

### Case fatality rate 

Use census as denominator

#### Over time

#### By geography

#### Spatio-temporal

### Risk factors for death

Matched case-control (or multinomial logistic regression) between:

1. In the field (asymptomatic, in the field, from xmal)
2. In the OPD (symptomatic, but quick to recover)
3. In the INPD (symptomatic, severe malaria)

Loop in Quique and Alfredo^[He looks at parasite differences that cause severe vs. asymptomatic]

```{r bib, include=FALSE}
# create a bib file for the R packages used in this document
knitr::write_bib(c('base', 'rmarkdown'), file = 'skeleton.bib')
```
