---
title: "Clinical malaria trends"
subtitle: "2002-2016"
author: ""
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, 
               echo = FALSE, 
               warning = FALSE, 
               message = FALSE, 
               error = TRUE, 
               cache = FALSE)
```

```{r}
library(tufte)
library(cism)
library(ggplot2)
library(ggthemes)  
library(cism)
library(tidyr)
library(RColorBrewer)

# Prepare data
source('prepare_data.R')
```

Quique Bassat, Joe Brew, Bea Galatas^[Collaborators list is non-exhaustive and alphabetical.]


# Analysis

## OPD

### OPD: Absolute number of malaria cases over time

The below chart shows the 

```{r}
x <- opd %>%
  mutate(month=format(date, "%m"),
         year=format(date, "%Y")) %>%
  group_by(year, month) %>%
  summarise(all=n(),
            malaria=length(which(malaria))) %>%
  mutate(p=malaria/all) %>%
  mutate(no_malaria = all - malaria) %>%
  mutate(dumdate=as.Date(paste0(year, "-", month, "-15")))

x_g <- gather(x, key, value, all:no_malaria)
x_g <- x_g %>%
         filter(key %in% c('malaria', 'no_malaria'))
x_g$key <- Hmisc::capitalize(gsub('_', ' ', x_g$key))
x_g$key <- factor(x_g$key, levels = rev(sort(unique(x_g$key))))

ggplot(x_g,
       aes(x = dumdate, 
           y = value,
           group = key,
           fill = key)) +
  geom_area(pos = 'stack', 
            alpha = 0.6,
            color = 'black',
            lwd = 0.3) +
  scale_fill_manual(name = '',
                    values = c('darkgreen', 'darkorange')) +
  labs(x = 'Date',
       y = 'Cases',
       title = 'Malaria and non-malaria cases over time',
       subtitle = 'OPD data') +
  theme_economist()
```

Clearly in the above chart, seasonality is driven by malaria. That said, even among non-malaria cases, there still appear to be seasonal trends.

```{r}
no_malaria <- x_g %>%
         filter(key == 'No malaria')
ggplot(data = no_malaria) +
  geom_area(aes(x = dumdate, y = value),
            fill = 'darkgreen', alpha= 0.6,
            lwd = 0.3,
            color = 'black') +
    labs(x = 'Date',
       y = 'Cases',
       title = 'Non-malaria cases over time',
       subtitle = 'OPD data') +
  theme_economist()
```

We can explore seasonality further by looking at each year.

```{r}
cols <- colorRampPalette(brewer.pal(n = 9, name = 'Spectral'))(length(unique(no_malaria$year)))
ggplot(data = no_malaria,
       aes(x = as.Date(paste0('1900-', as.numeric(month), '-01')),
           y = value,
           group = year,
           color = year)) +
  geom_line() +
  scale_color_manual(name = 'Year',
                     values = cols) +
  labs(x = 'Month',
       y = 'Cases',
       title = 'Non-malaria cases seasonality',
       subtitle = 'Each line is one year') +
  theme_economist() +
  scale_x_date(date_labels = "%b")

```

Malaria as a percentage of all cases is displayed in the below chart.

```{r}
ggplot(data = x,
       aes(x = dumdate,
           y = p * 100)) +
  geom_area(fill = 'darkorange', alpha = 0.6) +
  ylim(0, 100) +
  theme_economist() +
  labs(x = 'Date',
       y = 'Percentage',
       title = 'Malaria as a percentage of all cases',
       subtitle = 'OPD data')
```

As with non-malaria cases, we can explore seasonality in malaria cases by overlapping years on top of one another.

```{r}
cols <- colorRampPalette(brewer.pal(n = 9, name = 'Spectral'))(length(unique(x$year)))
ggplot(data = x,
       aes(x = as.Date(paste0('1900-', as.numeric(month), '-01')),
           y = p * 100,
           group = year,
           color = year)) +
  geom_line() +
  scale_color_manual(name = 'Year',
                     values = cols) +
  labs(x = 'Month',
       y = 'Percentage',
       title = 'Malaria cases seasonality',
       subtitle = 'Each line is one year') +
  theme_economist() +
  scale_x_date(date_labels = "%b")
```

The above charts aggregated data at the monthly level. An alternative is to examine at the daily level.

```{r}
x <- opd %>%
  group_by(date) %>%
  summarise(all=n(),
            malaria=length(which(malaria))) %>%
  mutate(p=malaria/all) %>%
  mutate(no_malaria = all - malaria) 

ggplot(data = x,
       aes(x = date,
           y = p * 100)) +
  geom_line(alpha = 0.6,
            color = 'darkorange') +
    labs(x = 'Date',
       y = 'Percentage',
       title = 'Malaria cases seasonality',
       subtitle = 'Daily variation') +
  theme_economist() 
```

Seasonality using daily data is best examined through faceting. The below chart shows malaria (as a percentage of all OPD cases) for each year separately.

```{r}
ggplot(data = x %>%
         mutate(year = as.numeric(format(date, '%Y'))),
       aes(x = date,
           y = p * 100)) +
  geom_point(alpha = 0.6,
            color = 'darkorange',
            size = 0.5) +
  facet_wrap(~year, scales = 'free_x') +
    theme_economist() +
  scale_x_date(date_labels = "%b") +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = 'Malaria as a percentage of all cases',
       subtitle = 'OPD',
       x = 'Date',
       y = 'Percentage')
```

To see trends from a more birds-eye view, we can also aggregate at the year level. Doing so results in the following.

```{r}
x <- opd %>%
  mutate(year = as.numeric(format(date, '%Y'))) %>%
  group_by(year) %>%
  summarise(all=n(),
            malaria=length(which(malaria))) %>%
  mutate(p=malaria/all) %>%
  mutate(no_malaria = all - malaria) 
ggplot(data = x,
       aes(x = year,
           y = malaria)) +
  geom_bar(stat = 'identity',
           fill = 'darkorange',
           alpha = 0.6) +
  labs(x = 'Year',
       y = 'Cases',
       title = 'Malaria cases by year',
       subtitle = 'OPD data') +
  theme_economist()
```

Instead of raw numbers (above), we can also examine yearly malaria in relative terms (below):

```{r}
x <- opd %>%
  mutate(year = as.numeric(format(date, '%Y'))) %>%
  group_by(year) %>%
  summarise(all=n(),
            malaria=length(which(malaria))) %>%
  mutate(p=malaria/all) %>%
  mutate(no_malaria = all - malaria) 
ggplot(data = x,
       aes(x = year,
           y = p * 100)) +
  geom_bar(stat = 'identity',
           fill = 'darkorange',
           alpha = 0.6) +
  labs(x = 'Year',
       y = 'Percentage',
       title = 'Malaria cases as percentage of all cases, by year',
       subtitle = 'OPD data') +
  theme_economist()
```


#### By age group

The distribution of all OPD visits is below.

```{r}
ggplot(data = opd,
       aes(x = age / 12)) +
  geom_density(fill = 'darkblue',
               alpha = 0.6) +
  labs(title = 'Distribution of OPD patients',
       subtitle = '0-15 years old only',
       x = 'Years',
       y = 'Density') +
  theme_economist()
```

For our analysis, we divide into 5 groups, with distributions as follows:

```{r}
ggplot(data = opd,
       aes(x = age_group)) +
  geom_bar(fill = 'darkblue',
               alpha = 0.6) +
  labs(title = 'Distribution of OPD patients',
       subtitle = '0-15 years old only',
       x = 'Years',
       y = 'Density') +
  theme_economist()
```

The total daily OPD caseload for each group is seen below. Each point is one day's caseload. The blue line is the smoothed trend.

```{r}
x <- opd %>%
  group_by(date, age_group) %>%
  tally

ggplot(data = x,
       aes(x = date, y = n)) +
  geom_point(alpha = 0.2,
             size = 0.3) +
  geom_smooth() +
  theme_economist() +
  labs(title = 'Total OPD cases',
       subtitle = 'By age group',
       x = 'Date',
       y = 'Cases') +
  facet_wrap(~age_group)
```

The below chart is identical to the above, but with malaria cases only.

```{r}
x <- opd %>%
  filter(malaria) %>%
  group_by(date, age_group) %>%
  tally

ggplot(data = x,
       aes(x = date, y = n)) +
  geom_point(alpha = 0.2,
             size = 0.3) +
  geom_smooth() +
  theme_economist() +
  labs(title = 'Malaria OPD cases',
       subtitle = 'By age group',
       x = 'Date',
       y = 'Cases') +
  facet_wrap(~age_group)
```

As an alternative to the above, we can examine each age group's malaria cases as a proportion of the total number of visists from patients of that age group.

```{r}
x <- opd %>%
  group_by(age_group, date) %>%
  summarise(malaria = length(which(malaria)),
            total = n()) %>%
  mutate(non_malaria = total - malaria,
         p = malaria / total)

cols <- colorRampPalette(brewer.pal(n = 9, name = 'Spectral'))(length(unique(x$age_group)))
ggplot(data = x,
       aes(x = date,
           y = p * 100,
           group = age_group,
           color = age_group)) +
  geom_line(alpha = 0.6) +
  scale_color_manual(name = '',
                     values = cols) +
  theme_economist() +
  labs(x = 'Date',
       y = 'Percentage',
       title = 'Malaria as percentage of all cases',
       subtitle = 'OPD, stratified by age group')
```


Since the above is to dense for clear understanding, we can facet for each year for a more in-depth exploration. The below chart shows one facet for each year, along with smoothed loess trend lines.

```{r}
ggplot(data = x %>%
         mutate(year = as.numeric(format(date, '%Y'))),
       aes(x = date,
           y = p * 100,
           group = age_group,
           color = age_group)) +
  geom_point(alpha = 0.6,
             size = 0.2) +
    stat_smooth(se = FALSE) +
  scale_color_manual(name = '',
                     values = cols) +
  theme_economist() +
  labs(x = 'Date',
       y = 'Percentage',
       title = 'Malaria as percentage of all cases',
       subtitle = 'OPD, stratified by age group') +
  facet_wrap(~year, scales = 'free_x') +
  scale_x_date(date_labels = "%b") 
```

It appears that seasonal risk is similar across age groups.

```{r}
x <- opd %>%
  mutate(month = as.numeric(format(date, '%m'))) %>%
  group_by(month, age_group) %>%
  summarise(malaria = length(which(malaria)),
            total = n()) %>%
  mutate(non_malaria = total - malaria,
         p = malaria / total)

cols <- colorRampPalette(brewer.pal(n = 9, name = 'Spectral'))(length(unique(x$age_group)))

ggplot(data = x,
       aes(x = month,
           y = p * 100)) +
  geom_area(stat = 'identity',
            color = 'black',
            lwd = 0.5,
            fill = 'darkorange',
            alpha = 0.6) +
  facet_wrap(~age_group, ncol = 1) +
  theme_economist() +
  labs(x = 'Month',
       y = 'Percentage',
       title = 'Age and seasonality',
       subtitle = 'Malaria cases as percentage of all OPD visits, by age')
```


### Incidence: Malaria cases over time (relative to population at risk)

### Clinical profile of malaria

*What does non-complicated malaria look like over time?*

- Parasite densitites^[We still need servolab stuff]
- Body temperature
- Hemoglobin
- Age distributions over time

#### Age group analysis

Cohort analysis with time-standardized x axis and cohort colored lines

(Everything up to here is for Katarina's paper)

### Hotspot identifical / Spatial scan statistic

Bayesian spatial-temporal analysis

Validate using actual locations.

Compare clinical hotspots from OPD with community hotspots from xmal.



## INPD trends

### Absolute cases over time

### Cases as percentage of all INPD cases over time

### Case fatality rate 

Use census as denominator

#### Over time

#### By geography

#### Spatio-temporal

### Risk factors for death

Matched case-control (or multinomial logistic regression) between:

1. In the field (asymptomatic, in the field, from xmal)
2. In the OPD (symptomatic, but quick to recover)
3. In the INPD (symptomatic, severe malaria)


```{r bib, include=FALSE}
# create a bib file for the R packages used in this document
knitr::write_bib(c('base', 'rmarkdown'), file = 'skeleton.bib')
```
